Stability in human interaction networks: sector relative sizes, prominence of topological measures and time activity statistics.

### 2015-05-07 paper.tex
incorporada revisao do chu

\section{Introduction}\label{sec:into}
Studies on human interaction networks have started long before modern computers, dating back to the nineteenth century, while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno~\cite{newmanBook}. With the increasing availability of data related to human interactions, research on these networks has grown continuously. Contributions can now be found in a variety of fields in the literature, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic. One of the approaches from an exact science perspective is to represent interaction networks as complex networks [ref.?][gmane,barabasiHumanDyn,newmanFriendship], with which 
several features of human interaction have been revealed. For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes. The dynamics of complex networks representing human interaction has also been addressed ~\cite{barabasiEvo,newmanEvolving}, but only to a limited extent, since research is normally focused on a particular metric or functionality, such as accessibility or community detection~\cite{access,newmanModularity}.

In this paper we analyze the evolution of human interaction networks, by considering interaction in email lists as their representative. Using a timeline of activity snapshots with a constant number of contiguous messages in email lists, we found a remarkable stability for several of the network properties. Because this stability was shared by all email lists, we advocate that some of the conclusions can be valid for more general classes of interaction networks. In particular, this allows us to discuss typologies in the context of such networks, in an attempt to bridge the gap between approaches based solely on data analysis (i.e. from a hard sciences perspective) and those relevant to the social sciences. This is important insofar as typologies are the canon of scientific literature for classification of human agents~\cite{typCanon}.

The paper is organized as follows. Section~\ref{sec:related} describes related work, while details of the data and methods of analysis are given in  Section~\ref{sec:data} and Section~\ref{sec:carac}. Section~\ref{sec:results} brings the results and discussion, leading to Section~\ref{sec:conc} for conclusions and further work.

\subsection{Related work}\label{sec:related}
Research on network evolution often considers solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Exceptions are reported in this section, with emphasis on those more closely related to the present article. Network types have been discussed with regard to the number of participants, intermittence of their activity and network longevity
~\cite{barabasiEvo}. Two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which can either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were reported with $\alpha=1$~\cite{bird} (as are web browsing and library loans~\cite{barabasiHumanDyn}), and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

Unreciprocated edges often exceed 50\% in the networks analyzed, which matches empirical evidence from the literature~\cite{newmanEvolving}. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were not considered in our study. Gender related behavior in mobile phone datasets was indeed reported~\cite{barabasiSex}, but this was not considered in the present article because email messages and addresses have no gender related metadata~\cite{GMANE}.

@@

We analyzed many email lists, but selected only four in order to make a thorough analysis, from which general properties can be inferred. These lists, selected as representing both a diverse set and ordinary lists, are:
\begin{itemize}
	\item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}, with participants holding hybrid artistic and technological interests, from different countries. English is the language used the most. Abbreviated as LAU from now on.
	\item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}, with participants from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
	\item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}, with computer programmers from different countries. English is the language used the most. Abbreviated as CPP from now on.
	\item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}, with Brazilian activists holding digital culture interests. Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
	\centering
	\caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
		$N$ is the number of participants (number of different email addresses).
		$\Gamma$ is the number of threads (count of messages without antecedent).
		$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
	A relation holds for all lists carefully considered: as the number of participants increases, the number of threads decreases.}
	\label{tab:genLists}
	\begin{tabular}{|l|c|c|c|c|c|}\hline
		list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
		\input{tables/tab1Geral}
	\end{tabular}
\end{table}

\section{Characterization methods}\label{sec:carac}
The email lists and the networks generated from them were characterized by using five procedures, namely: 1) statistics of activity along time, from seconds to years; 2) sectioning of the networks in hubs, intermediary and peripheral vertices; 3) topological metrics and their dispersion; 4) iterative visualization and data mining; 5) typological speculation about networks and participants.
Each of these procedures are described below.

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. This resulted in histograms from which patterns could be drawn. The ratio $\frac{b_h}{b_l}$ between the highest and lowest incidences on the histograms served as a hint of how the observed distribution is compared to a uniform distribution.

@@

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted, both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this article are directed and weighted, the more informative of trivial possibilities, i.e. we did not investigate directed unweighted, undirected weighted, and undirected unweighted representations of the interaction networks. 
The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to the edge weight. Self-loops were regarded as non-informative and discarded. These human social interaction networks are reported in the literature as exhibiting scale-free and small world properties, as expected for (some) social networks~\cite{bird,newmanBook}.

@@

Edges can be created from all antecedent message authors on the message-response thread to each message author.
We only linked the immediate antecedent to the new message author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a weaker connection between $x$ and $z$. Potential interpretations for this weaker connection are: double length, half weight or with one more ``obstacles''. This suggests the adequacy of centrality measurements to account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

@@

The lower degree fat tail represents the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is: 1) vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in scale-free phenomena, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser_}
	\caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
		intermediary vertices, while the former has more peripheral and hub vertices. The sector borders are defined by the two intersections $k_\Leftarrow$ and $k_\Rightarrow$ of the connectivity distributions. Characteristic degrees
																											are in compact intervals of degree: $[0,k_\Leftarrow]$, $(k_\Leftarrow,k_\Rightarrow]$, $(k_\Rightarrow,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
																											\label{fig:setores}
																										\end{figure}

																										To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer with which there are at least $\eta$ vertices with degree larger than or equal $k_i$, and less than or equal $k_{j}$. This changes equation~\ref{criterio} to:

@@

																										In other words, let $\gamma$ and $\phi$ be integers in the intervals $1 \leq \gamma \leq 6$, $1 \leq \phi \leq 3$, and the basic six Erd\"os sectioning possibilities $\{E_{\gamma}\}$ have three Erd\"os sectors $E_{\gamma}= \{e_{\gamma, \phi} \}$ defined as:

@@

																										\begin{itemize}
																											\item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified (usually) do not reach 100\%, which is indicated by a black line in Figures~\ref{fig:sectIL}.
																											\item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics. Therefore, a vertex may belong to more than one class, and total members may add more than 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
																											\item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
																											\item Inclusivist cascade $C_4$: vertices are hubs if they are classified as so according to any of the metrics. The remaining vertices are classified as intermediary if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
																											\item Exclusivist externals $C_5$: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes by any metric. The rest of the nodes are classified as intermediary.
																											\item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
																										\end{itemize}

@@

																										The simplification of all the compound possibilities to the small set listed above can be formalized in strict mathematical terms, but this was considered out of the scope for current interests. It is worth noting that the exclusivist cascade is the same sectioning of an inclusivist cascade from periphery to hubs, but with inverted order of sectors precedence. These compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Support Information.

@@

																										\subsection{Evolution and visualization of the networks}\label{sec:viz}
																										The evolution of the networks was observed within a fixed number of messages, which we refer to as the window size $ws$. This same number of contiguous messages $ws$ was considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used both to perform the Erd\"os sectioning and apply PCA for the topological metrics.  
																										The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics. Such changes could be visualized with the tools described below.

																										Networks were visualized with animations, image galleries and online gadgets developed specifically for this research~\cite{animacoes,galGMANE,appGMANE}. Such visualizations were crucial to guide research into the most important features of network evolution, and prompted us to capture the prominence of topological metrics along time using mean and standard deviations. Furthermore, the size of three sectors could be visualized in a timeline fashion. Visualization of network structure was especially useful as part of the email lists data mining, from which parts of relevant structures and results were driven.

																										%\subsection{Typology speculation}\label{subsec:typ}
																										%Qualitative, typological speculations were the result of all the methodology. More specifically, the Erd\"os sectors was regarded as yielding a \emph{primitive typology} of human agents in social contexts with at least dozens of participants. 
																										%Visualization and data mining were efficient in fueling speculations about the qualities of each type. 
																										%The stability found in principal components of topological measures suggested that this primitive typology is general for (practically) all human contexts.
																										%Also, the first author is a developer with cultural interests and has been directly and indirectly part of these communities for more than a decade.  This gave further safety about assumptions although care was taken not to regard this acquaintanceship as primary source or as definitive evidence.
																										%The interested reader should see Appendix~\ref{ap:typ} for further context, 
																										%as it might be considered audacious to bridge from a physics-based, 
																										%quantitative classification to a qualitative and speculative approach.

																										\section{Results and discussion}\label{sec:results}

																										%Remarkable features from the analysis of the four email lists are:
																										%\begin{itemize}
																										%    \item The activity along time is practically the same for all lists, thus suggesting stable patterns.
																										%    \item The fraction of participants in each Erd\"os sector is stable along time and can be determined even with very few messages 
																										%    \item The topological metrics combine into principal components in PCA in the same way for all lists and all snapshots (). 
																										%        \item Symmetry measures of the topology, as defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
																										%    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
																										%\end{itemize}

																										\subsection{Activity along time}\label{constDisc}

																										The activity along time, in terms of seconds, minutes, hours, days and months,  is practically the same for all lists.
																										Using circular statistics we calculated average values and their dispersion for activity in all time scales, and found that both average and dispersion were very similar in all the lists. We chose to give detailed values in Table~\ref{tab:circ}-\ref{tab:min2} because these numbers can actually be used for characterizing nodes (participants) in other networks, as they are independent of the network under analysis. For example, they may serve for identification of outliers in a community.

																									In the scale of seconds and minutes, activity obeys a homogeneous pattern, with the messages being slightly more evenly distributed in all lists than in simulations\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.} using uniform distribution: $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
																									Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected. Circular dispersion is maximized and the mean has little meaning as indicated in Table~\ref{tab:circ}. As for the hours of the day, an abrupt peak appeared around 11am with the most active period being the afternoon. Days of the week revealed a decrease of at least one third and at most two thirds of activity on weekends. Days of the month were regarded as homogeneous with an inconclusive slight tendency of the first week being more active. Months of the year revealed patterns matching usual work and academic calendars. The time period examined here was not sufficient for the analysis of activity along the years. These patterns are exemplified in Tables~\ref{tab:hin}-\ref{tab:min2}.

@@

																									\FloatBarrier
																									\begin{table}
																										\caption{Activity percentages along the days of the week for the four email lists. Higher activity was observed during weekdays, with a decrease of activity on weekends of at least one third and two thirds in extreme cases.}
																										\begin{center}
																											\begin{tabular}{ | l |  c | c | c | c | c |   c | c |}
																												\hline
																												& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
																												\input{tables/tabWeekdays}
																											\end{tabular}
																										\end{center}
																										\label{tab:win}
																									\end{table}
																									\begin{table}
																										\caption{Activity in the days along the month for MET list. Nearly identical distributions are found on other lists as indicated in the Supporting Information. Although slightly higher activity rates are found in the beginning of the month, the most important feature seems to be the homogeneity made explicit by the high circular dispersion in Table~\ref{tab:circ}.}
																										\footnotesize
																										\input{tables/tabMonthdaysMET}
																										\label{tab:min}
																									\end{table}

@@

																									%\vspace*{.5cm}
																									\subsection{Scalable fat-tail structure: constancy of membership fractions in the Erd\"os sectors}\label{subsec:pih}

																									The distribution of vertices in the hubs, intermediary, periphery Erd\"os sectors is remarkably stable along time, provided that a sufficiently large sample of 200 or more messages is considered. 
																									Moreover, the same distribution applies to the networks of all the four email lists, as demonstrated in Figure~\ref{fig:sectIL} and the Supporting Information. 
																									Typically, $\approx [3-12]\%$ of the vertices are found to be hubs, $\approx [15-45]\%$ are intermediary and $\approx [44-81]\%$ are peripheral, which is consistent with the literature~\cite{secFree}.
																									These results hold for the total, in and out degrees and strengths.
																									Stable distributions can also be obtained for 100 or less messages if classification of the three sectors is performed with one of the compound criteria established in Section~\ref{sectioning}. 
																									The networks hold their basic structure with as few as 10-50 messages; concentration of activity and the abundance of low-activity participants take place even with very few messages, which is highlighted in the last figures of the Supporting Information.
																									A minimum window size for observation of more general properties might be inferred by monitoring the giant component and the degeneration of the Erd\"os sectors.

@@

																									Activity is highly concentrated on the hubs, while a very large number of peripheral vertices contribute to only a fraction of the activity. This is expected for a system with scale-free behavior, as confirmed with the data in Table~\ref{autores} for the distribution of activity among participants. (certo?)

@@

																										\begin{table}[!h]
																											\caption{Loadings for the 14 metrics into the principal components for the MET list, $ws=1000$ messages in 20 disjoint positioning. The clustering coefficient (cc) appears as the first metric in the Table, followed by 7 centrality metrics and 6 symmetry-related metrics. Note that the centrality measurements, including degrees, strength and betweenness centrality, are the most important contributors for the first principal component, while the second component is dominated by symmetry metrics. The clustering coefficient is only relevant for the third principal component. The three components have in average 80.36\% of the variance.}
																												\footnotesize
																												\input{tables/tabPCA3CPP}
																												\label{tab:pcain}
																											\end{table}

>>>>

\section{Introduction}\label{sec:into}
Studies on human interaction networks have started long before modern computers, dating back to the nineteenth century, while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno in mid twentieth century~\cite{newmanBook}. With the increasing availability of data related to human interactions, research on these networks has grown continuously. Contributions can now be found in a variety of fields, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic. One of the approaches from an exact science perspective is to represent interaction networks as complex networks~\cite{gmane,barabasiHumanDyn,newmanFriendship}, with which 
several features of human interaction have been revealed. For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes. The dynamics of complex networks representing human interaction has also been addressed ~\cite{barabasiEvo,newmanEvolving}, but only to a limited extent, since research is normally focused on a particular metric or functionality, such as accessibility or community detection~\cite{access,newmanModularity}.

In this paper we analyze the evolution of human interaction networks, by considering interaction in email lists as their representative. Using a timeline of activity snapshots with a constant number of contiguous messages in email lists, we found a remarkable stability for several of the network properties. Because this stability was shared by all email lists, is coherent with literature, and has been observed in some networks from Twitter and Facebook, we advocate that some of the conclusions might be valid for more general classes of interaction networks. In particular, this allows us to discuss typologies in the context of such networks, in an attempt to bridge the gap between approaches based solely on data analysis (i.e. from a hard sciences perspective) and those relevant to the social sciences. This is important insofar as typologies are the canon of scientific literature for classification of human agents~\cite{typCanon}.

The paper is organized as follows. Section~\ref{sec:related} describes related work, while details of the data and methods of analysis are given in  Section~\ref{sec:data} and Section~\ref{sec:carac}. Section~\ref{sec:results} brings the results and discussion, leading to Section~\ref{sec:conc} for conclusions.

\subsection{Related work}\label{sec:related}
Research on network evolution often considers solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Exceptions are reported in this section, with emphasis on those more closely related to the present article.

Network types have been discussed with regard to the number of participants, intermittence of their activity and network longevity~\cite{barabasiEvo}. Two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which can either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were reported with $\alpha=1$~\cite{bird} (as are web browsing and library loans~\cite{barabasiHumanDyn}), and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

Unreciprocated edges often exceed 50\% in the networks analyzed, which matches empirical evidence from the literature~\cite{newmanEvolving} and motivated the inclusion of symmetry metrics on our analysis. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were not considered in our study. Gender related behavior in mobile phone datasets was indeed reported~\cite{barabasiSex}, but this was not considered in the present article because email messages and addresses have no gender related metadata~\cite{GMANE}.

@@

We analyzed many email lists (and data from Facebook and Twitter) but selected only four in order to make a thorough analysis, from which general properties can be inferred. These lists, selected as representing both a diverse set and ordinary lists, are:
\begin{itemize}
	\item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}, with participants holding hybrid artistic and technological interests, from different countries. English is the language used the most. Abbreviated as LAU from now on.
	\item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}, with participants from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
	\item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}, with computer programmers from different countries. English is the language used the most. Abbreviated as CPP from now on.
	\item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}, with Brazilian activists holding digital culture interests. Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
	\centering
	\caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
		$N$ is the number of participants (number of different email addresses).
		$\Gamma$ is the number of threads (count of messages without antecedent).
		$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
	A relation holds for all lists carefully considered: within a same number of messages, as the number of participants increases, the number of threads decreases.}
	\label{tab:genLists}
	\begin{tabular}{|l|c|c|c|c|c|}\hline
		list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
		\input{tables/tab1Geral}
	\end{tabular}
\end{table}

\section{Characterization methods}\label{sec:carac}
The email lists and the networks generated from them were characterized using five procedures, namely: 1) statistics of activity along time, in scales from seconds to years; 2) sectioning of the networks in hubs, intermediary and peripheral vertices; 3) dispersion of basic topological metrics; 4) iterative visualization and data inspection.
Each of these procedures are described below.

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. This resulted in histograms from which patterns could be drawn. The ratio $\frac{b_h}{b_l}$ between the highest and lowest incidences on the histograms served as a hint of how close the observed distribution is to a uniform distribution.

@@

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted, both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this article are directed and weighted, the more informative of trivial possibilities, i.e. we did not investigate directed unweighted, undirected weighted, and undirected unweighted representations of the interaction networks. 
The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to the edge weight. Selfloops were regarded as non-informative and discarded. These human social interaction networks are reported in the literature as exhibiting scale-free and small world properties, as expected for (some) social networks~\cite{bird,newmanBook}.

@@

Edges can be created from all antecedent message authors on the message-response thread to each message author.
We only linked the immediate antecedent to the new message author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a weaker connection between $x$ and $z$. Potential interpretations for this weaker connection are: double length, half weight or with one more ``obstacles''. This suggests the adequacy of centrality measurements to account for the connectivity with all nodes, such as betweenness centrality, eigenvector centrality and accessibility~\cite{luMeasures,access}.

@@

The lower degree fat tail represents the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is: 1) vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in the real network, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser_}
	\caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks.
		The latter has more intermediary vertices, while the former has more peripheral and hub vertices.
		The sector borders are defined by the two intersections $k_L$ and $k_R$ of the connectivity distributions.
		Characteristic degrees are in the compact intervals: $[0,k_L]$, $(k_L,k_R]$, $(k_R,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
		\label{fig:setores}
\end{figure}
i	To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer with which there are at least $\eta$ vertices with degree larger than or equal $k_i$, and less than or equal $k_{j}$. This changes equation~\ref{criterio} to:

@@

In other words, let $\gamma$ and $\phi$ be integers in the intervals $1 \leq \gamma \leq 6$, $1 \leq \phi \leq 3$, and each of the basic six Erd\"os sectioning possibilities $\{E_{\gamma}\}$ have three Erd\"os sectors $E_{\gamma}= \{e_{\gamma, \phi} \}$ defined as:

@@

\begin{itemize}
	\item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified (usually) do not reach 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics. Therefore, a vertex may belong to more than one class, and total members may add more than 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
	\item Inclusivist cascade $C_4$: vertices are hubs if they are classified as so according to any of the metrics. The remaining vertices are classified as intermediary if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
	\item Exclusivist externals $C_5$: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes by any metric. The rest of the nodes are classified as intermediary.
	\item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
\end{itemize}

@@

The simplification of all the compound possibilities to the small set listed above can be formalized in strict mathematical terms, but this was considered out of the scope for current interests. It is worth noting that the exclusivist cascade is the same sectioning of an inclusivist cascade from periphery to hubs, but with inverted order of sectors precedence. These compound criteria can be used to examine network sections considering all degrees and strengths and are specially useful in the case of a low number of messages, such as in the last figures of Support Information.

@@

\subsection{Evolution and visualization of the networks}\label{sec:viz}
The evolution of the networks was observed within a fixed number of messages, which we refer to as the window size $ws$. This same number of contiguous messages $ws$ was considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used both to perform the Erd\"os sectioning and to apply PCA for the topological metrics.  
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics, which is exhibited in the figures of Support Information.

Networks were visualized with animations, image galleries and online gadgets developed specifically for this research~\cite{animacoes,galGMANE,appGMANE}. Such visualizations were crucial to guide research into the most important features of network evolution. Furthermore, the size of the three Erd\"os sectors could be visualized in a timeline fashion. Visualization of network structure was especially useful in the inspection of data and derived structures from the email lists.

\section{Results and discussion}\label{sec:results}
%Remarkable features from the analysis of the four email lists are:
%\begin{itemize}
%    \item The activity along time is practically the same for all lists, thus suggesting stable patterns.
%    \item The fraction of participants in each Erd\"os sector is stable along time and can be determined even with very few messages 
%    \item The topological metrics combine into principal components in PCA in the same way for all lists and all snapshots (). 
%        \item Symmetry measures of the topology, as defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
%    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
%\end{itemize}
\subsection{Activity along time}\label{constDisc}
The activity along time, in terms of seconds, minutes, hours, days and months,  is practically the same for all lists.
Histograms in each time scale were calculated as were circular average values and their dispersion.
We chose to give detailed values in Table~\ref{tab:circ}-\ref{tab:min2} because these numbers can actually be used for characterizing nodes (participants) in other networks, and networks themselves, as they are independent of the network under analysis. For example, they may serve for identification of outliers in a community.

In the scale of seconds and minutes, activity obeys a homogeneous pattern, with the messages being slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}. In the networks, $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$ while simulations reach these values but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected. Circular dispersion is maximized and the mean has little meaning as indicated in Table~\ref{tab:circ}. As for the hours of the day, an abrupt peak appeared around 11am with the most active period being the afternoon. Days of the week revealed a decrease of at least one third and at most two thirds of activity on weekends. Days of the month were regarded as homogeneous with an inconclusive slight tendency of the first week being more active. Months of the year revealed patterns matching usual work and academic calendars. The time period examined here was not sufficient for the analysis of activity along the years. These patterns are exemplified in Tables~\ref{tab:hin}-\ref{tab:min2}.

@@

\FloatBarrier
\begin{table}
	\caption{Activity percentages along the days of the week for the four email lists. Higher activity was observed during weekdays, with a decrease of activity on weekends of at least one third and two thirds in extreme cases.}
	\begin{center}
		\begin{tabular}{ | l |  c | c | c | c | c |   c | c |}
			\hline
			& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
			\input{tables/tabWeekdays}
		\end{tabular}
	\end{center}
	\label{tab:win}
\end{table}

\begin{table}
	\caption{Activity in the days along the month for MET list. Nearly identical distributions are found on other lists as indicated in the Supporting Information. Although slightly higher activity rates are found in the beginning of the month, the most important feature seems to be the homogeneity made explicit by the high circular dispersion in Table~\ref{tab:circ}.}
	\footnotesize
	\input{tables/tabMonthdaysMET}
	\label{tab:min}
\end{table}

@@

\subsection{Scalable fat-tail structure: constancy of membership fractions in the Erd\"os sectors}\label{subsec:pih}

The distribution of vertices in the hubs, intermediary, periphery Erd\"os sectors is remarkably stable along time, provided that a sufficiently large sample of 200 or more messages is considered. 
Moreover, the same distribution applies to the networks of all the four email lists, as demonstrated in Figure~\ref{fig:sectIL} and the Supporting Information. 
Activity is highly concentrated on the hubs, while a very large number of peripheral vertices contribute to only a fraction of the activity. This is expected for a system with a scale-free trace, as confirmed with the data in Table~\ref{autores} for the distribution of activity among participants.

Typically, $\approx [3-12]\%$ of the vertices are found to be hubs, $\approx [15-45]\%$ are intermediary and $\approx [44-81]\%$ are peripheral, which is consistent with the literature~\cite{secFree}.
These results hold for the total, in and out degrees and strengths.
Stable distributions can also be obtained for 100 or less messages if classification of the three sectors is performed with one of the compound criteria established in Section~\ref{sectioning}. 
The networks hold their basic structure with as few as 10-50 messages; concentration of activity and the abundance of low-activity participants take place even with very few messages, which is highlighted in the last figures of the Supporting Information.
A minimum window size for observation of more general properties might be inferred by monitoring the giant component and the degeneration of the Erd\"os sectors.

@@

@@

\begin{table}[!h]
	\caption{Loadings for the 14 metrics into the principal components for the MET list, $ws=1000$ messages in 20 disjoint positioning. The clustering coefficient (cc) appears as the first metric in the Table, followed by 7 centrality metrics and 6 symmetry-related metrics. Note that the centrality measurements, including degrees, strength and betweenness centrality, are the most important contributors for the first principal component, while the second component is dominated by symmetry metrics. The clustering coefficient is only relevant for the third principal component. The three components have in average 80\% of the variance.}
		\footnotesize
		\input{tables/tabPCA3CPP}
		\label{tab:pcain}
	\end{table}
